---
title: "Experimental Data Cleaning and Checking Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Experimental Data Cleaning and Checking Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

The `excheckr` package provides tools for cleaning and validating experimental data. This vignette demonstrates a complete workflow for a randomized experiment, showing how to:

1. Inspect and visualize covariate missingness
2. Generate and apply imputation code
3. Create outcome missingness indicators
4. Check for balance across treatment conditions
5. Check for differential attrition

## Setup

```{r setup}
library(excheckr)
library(tidyverse)
library(estimatr)

set.seed(2024)
```

# Simulating Experimental Data

Let's create a realistic experimental dataset with:
- Treatment assignment (`Z`)
- Covariates with some missingness (`X_*`)
- Outcomes with differential missingness (`Y_*`)

```{r simulate-data}
n <- 500

# Treatment assignment (randomized)
dat <- data.frame(
  Z = rbinom(n, 1, 0.5),
  cluster_id = sample(1:25, n, replace = TRUE)
)

# Covariates with realistic missingness patterns
dat <- dat |>
  mutate(
    # Demographics
    X_age = rnorm(n, 45, 15),
    X_gender = sample(c("Male", "Female", "Other"), n, replace = TRUE, 
                      prob = c(0.48, 0.48, 0.04)),
    X_income = exp(rnorm(n, 11, 0.8)) * 1000,
    
    # Political variables
    X_party = factor(sample(c("Democrat", "Republican", "Independent"), n, 
                           replace = TRUE, prob = c(0.35, 0.30, 0.35))),
    X_ideology = sample(1:7, n, replace = TRUE),
    
    # Education
    X_education = factor(sample(c("HS", "Some College", "BA", "Graduate"), n,
                               replace = TRUE, prob = c(0.25, 0.30, 0.30, 0.15)))
  )

# Introduce MCAR missingness in covariates
dat$X_age[sample(1:n, 25)] <- NA
dat$X_income[sample(1:n, 40)] <- NA
dat$X_party[sample(1:n, 30)] <- NA
dat$X_ideology[sample(1:n, 35)] <- NA

# Outcomes
# Y_attitude: treatment increases missingness slightly (differential attrition)
# Y_behavior: no differential attrition
dat <- dat |>
  mutate(
    Y_attitude_latent = 3 + 0.5 * Z + 0.01 * X_age + rnorm(n),
    Y_behavior_latent = 0.3 + 0.2 * Z + rnorm(n, 0, 0.5)
  )

# Create missingness in outcomes
# Y_attitude: 25% missing in control, 35% missing in treatment (differential!)
miss_attitude <- rbinom(n, 1, ifelse(dat$Z == 1, 0.35, 0.25))
dat$Y_attitude <- ifelse(miss_attitude == 1, NA, dat$Y_attitude_latent)

# Y_behavior: 20% missing in both groups (no differential attrition)
miss_behavior <- rbinom(n, 1, 0.20)
dat$Y_behavior <- ifelse(miss_behavior == 1, NA, dat$Y_behavior_latent)

# Remove latent variables
dat <- dat |> select(-Y_attitude_latent, -Y_behavior_latent)

# View first few rows
head(dat)
```

# Step 1: Inspect Covariate Missingness

Before imputing, we should understand the patterns of missingness in our covariates.

```{r check-missingness}
# Check missingness patterns
check_covariate_missingness(dat)
```

The summary shows:
- `X_income` has the most missingness (8%)
- `X_ideology` and `X_party` also have notable missingness
- The heatmap shows joint missingness patterns

The diagonal shows univariate missingness, while off-diagonal cells show cases missing both variables. High off-diagonal values suggest missingness may not be completely at random.

# Step 2: Generate and Apply Imputation Code

Now we'll generate code to impute missing covariates. The `write_*_code()` functions print code that you can review and customize before running.

```{r generate-imputation-code}
# Generate imputation code for all X_ variables
write_covariate_imputation_code(dat)
```

This generates tidyverse code that:
- Imputes numeric variables with their median
- Imputes categorical variables with their mode
- Creates missingness indicator variables (`_missing`)
- Creates imputed versions (`_nona`)

Now let's apply this imputation:

```{r apply-imputation}
dat <-
  dat |>
  mutate(
    X_age_nona = replace_na(X_age, median(X_age, na.rm = TRUE)),
    X_age_missing = if_else(is.na(X_age), 1, 0),
    X_gender_nona = replace_na(X_gender, stat_mode(X_gender)),
    X_gender_missing = if_else(is.na(X_gender), 1, 0),
    X_income_nona = replace_na(X_income, median(X_income, na.rm = TRUE)),
    X_income_missing = if_else(is.na(X_income), 1, 0),
    X_party_nona = replace_na(X_party, stat_mode(X_party)),
    X_party_missing = if_else(is.na(X_party), 1, 0),
    X_ideology_nona = replace_na(X_ideology, median(X_ideology, na.rm = TRUE)),
    X_ideology_missing = if_else(is.na(X_ideology), 1, 0),
    X_education_nona = replace_na(X_education, stat_mode(X_education)),
    X_education_missing = if_else(is.na(X_education), 1, 0)
  )

# Verify no missingness in imputed variables
dat |>
  summarize(
    across(ends_with("_nona"), ~sum(is.na(.)))
  )
```

Perfect! All imputed variables have zero missing values.

# Step 3: Create Outcome Missingness Indicators

For outcomes, we typically don't impute. Instead, we create missingness indicators to check for differential attrition.

```{r generate-outcome-missing}
# Generate code for outcome missingness indicators
write_outcome_missingness_dummies_code(dat)
```

Apply this code:

```{r apply-outcome-missing}
dat <- 
  dat |>
  mutate(
    Y_attitude_missing = if_else(is.na(Y_attitude), 1, 0),
    Y_behavior_missing = if_else(is.na(Y_behavior), 1, 0)
  )

# Check missingness rates by treatment
dat |>
  group_by(Z) |>
  summarize(
    n = n(),
    attitude_missing_rate = mean(Y_attitude_missing),
    behavior_missing_rate = mean(Y_behavior_missing)
  )
```

We can see differential missingness in `Y_attitude` (35% in treatment vs 25% in control) but not in `Y_behavior` (20% in both groups).

# Step 4: Check Balance

Before analyzing treatment effects, we should verify that randomization successfully balanced covariates across treatment conditions.

```{r check-balance}
# Check balance on imputed covariates
balance_results <- check_balance(
  dat, 
  treatment = Z,
  covariates = c("X_age_nona", "X_income_nona", "X_ideology_nona", 
                 "X_gender", "X_party_nona", "X_education"),
  clusters = cluster_id,
  se_type = "CR2"
)
```

The results show the coefficient from regressing each covariate on treatment. For categorical variables, you'll see multiple rows (one per level, excluding the reference category).

Key things to look for:
- Coefficients close to zero (treatment doesn't predict covariates)
- Non-significant p-values
- For categorical variables, check each level

We can also generate code to do this manually:

```{r generate-balance-code}
write_balance_check_code(
  dat,
  treatment = Z,
  covariates = c("X_age_nona", "X_income_nona"),
  clusters = cluster_id,
  se_type = "CR2"
)
```

# Step 5: Check for Differential Attrition

Differential attrition occurs when treatment affects the probability of outcome missingness. This can bias treatment effect estimates.

```{r check-attrition}
# Check if treatment predicts outcome missingness
attrition_results <- check_attrition(
  dat,
  treatment = Z,
  clusters = cluster_id,
  se_type = "CR2"
)
```

The results show:
- **Y_attitude_missing**: Significant positive coefficient (~0.10, p < 0.05), indicating differential attrition. Treatment increases missingness by about 10 percentage points.
- **Y_behavior_missing**: Non-significant coefficient, indicating no differential attrition.

We can also generate the code:

```{r generate-attrition-code}
write_attrition_check_code(
  dat,
  treatment = Z,
  clusters = cluster_id,
  se_type = "CR2"
)
```

# Step 6: Check Missingness Indicator Balance

A more stringent check: are the missingness indicators themselves balanced? This tests whether treatment affects who is missing on covariates.

```{r check-missing-balance}
# Check if treatment predicts covariate missingness
missing_balance <- check_balance(
  dat,
  treatment = Z,
  covariates = c("X_age_missing", "X_income_missing", 
                 "X_party_missing", "X_ideology_missing"),
  clusters = cluster_id,
  se_type = "CR2"
)
```

If these are balanced, it suggests that missingness in covariates is not related to treatment assignment, which supports the assumption that our imputation approach is valid.

# Complete Workflow Summary

Here's the recommended workflow for experimental data:

## 1. Initial Data Inspection
```r
check_covariate_missingness(dat)
```

## 2. Impute Covariates
```r
# Generate code
write_covariate_imputation_code(dat)

# Review and apply the generated code
dat <- dat |>
  mutate(
    # ... paste generated code here ...
  )
```

## 3. Create Outcome Missingness Indicators
```r
# Generate code
write_outcome_missingness_dummies_code(dat)

# Apply code
dat <- dat |>
  mutate(
    # ... paste generated code here ...
  )
```

## 4. Validate the Experiment
```r
# Check covariate balance
check_balance(dat, Z)

# Check for differential attrition
check_attrition(dat, Z)

# Check missingness indicator balance (optional but recommended)
check_balance(dat, Z, covariates = ends_with("_missing"))
```

## 5. Address Issues
- **Imbalanced covariates**: Consider covariate adjustment in your analysis
- **Differential attrition**: Report as a limitation; consider bounds analysis or weighting
- **Imbalanced missingness indicators**: May suggest MNAR (missing not at random); consider sensitivity analyses

# Conclusion

The `excheckr` package streamlines the process of cleaning and validating experimental data by:

1. **Generating reproducible code** with `write_*_code()` functions that you can review before applying
2. **Running diagnostic checks** with `check_*()` functions that produce tidy, publication-ready results
3. **Following best practices** for handling missingness in experimental data

This workflow helps ensure your experimental analysis is:
- **Transparent**: All data transformations are explicit
- **Reproducible**: Generated code can be saved and reused
- **Valid**: Balance and attrition checks catch potential issues

For more information on individual functions, see their help pages (e.g., `?check_balance`).
