---
title: "Checking One Experiment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Checking One Experiment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
options(width = 100)
```

```{r setup, message = FALSE}
library(excheckr)
library(dplyr)
library(tidyr)
library(randomizr)
```

A cleaned experimental dataset invites four questions. Are the outcomes on the
scale you think they are on? Which covariates have missing values, and what did
you do about them? Did randomization balance the covariates? Did treatment affect
who is missing an outcome? This vignette answers all four on a simulated study
built with known problems, so that the checks have something to find.

Every check takes a `study_id` and appends it to its output. That is what makes
the results stackable across studies later, which is the subject of
`vignette("checking_many_studies")`.

## A study with planted problems

Treatment is assigned at the cluster level, which matters later: a clustered
design changes which balance test is trustworthy. `X_income` and `X_party` carry
missing values, and `Y_attitude` is missing more often among the treated, so the
attrition check should flag it and the balance checks should not.

```{r simulate}
set.seed(123)

n_clusters <- 25
cluster_size <- 20
n <- n_clusters * cluster_size

dat <- tibble(
  cluster_id = rep(seq_len(n_clusters), each = cluster_size),
  X_age = rnorm(n, 45, 15),
  X_gender = sample(c("Male", "Female", "Other"), n, replace = TRUE,
                    prob = c(0.48, 0.48, 0.04)),
  X_income = exp(rnorm(n, 11, 0.8)) * 1000,
  X_party = factor(sample(c("Democrat", "Republican", "Independent"), n,
                          replace = TRUE, prob = c(0.35, 0.30, 0.35))),
  X_ideology = sample(1:7, n, replace = TRUE)
) |>
  mutate(
    Z = cluster_ra(clusters = cluster_id),
    # Covariates go missing at random, unrelated to treatment
    X_income = if_else(runif(n) < 0.08, NA_real_, X_income),
    X_party = if_else(runif(n) < 0.06, NA, X_party),
    # Y_attitude drops out more often among the treated; Y_behavior does not
    Y_attitude = if_else(runif(n) < if_else(Z == 1, 0.35, 0.25), NA_real_,
                         plogis(0.5 * Z + 0.01 * X_age + rnorm(n))),
    Y_behavior = if_else(runif(n) < 0.20, NA_real_,
                         plogis(0.2 * Z + rnorm(n, 0, 0.5)))
  )

dat
```

## Are the outcomes on the expected scale?

`check_y_bounds()` reports the range of every `Y_` column and flags anything
outside the unit interval. Both outcomes here are probabilities, so both should
pass.

```{r ybounds}
check_y_bounds(dat, study_id = "demo_study_1")
```

A `FALSE` in `in_bounds` almost always means a cleaning error rather than a
finding: a percentage left on a 0 to 100 scale, or a reverse-coded item that
never got recoded. Fix it at the source.

## Which covariates are missing, and were they imputed?

`check_covariate_missingness()` prints a summary and returns a joint-missingness
heatmap. The diagonal is each covariate's own missingness rate; an off-diagonal
cell is the share of rows missing both, so a cell close to its diagonal
neighbours means the two go missing together.

```{r missingness}
missingness <- check_covariate_missingness(dat)
```

`write_covariate_imputation_code()` does not impute anything. It prints code for
you to read and paste, so the imputation lands in your script where a reader can
see it rather than inside a function call.

```{r generate-imputation}
write_covariate_imputation_code(dat)
```

Numeric covariates get the median, categorical ones the mode via `stat_mode()`,
and each gets a `_missing` indicator alongside the imputed `_nona` version.
Pasting that output gives:

```{r apply-imputation}
dat <-
  dat |>
  mutate(
    X_age_nona = replace_na(X_age, median(X_age, na.rm = TRUE)),
    X_age_missing = if_else(is.na(X_age), 1, 0),
    X_gender_nona = replace_na(X_gender, stat_mode(X_gender)),
    X_gender_missing = if_else(is.na(X_gender), 1, 0),
    X_income_nona = replace_na(X_income, median(X_income, na.rm = TRUE)),
    X_income_missing = if_else(is.na(X_income), 1, 0),
    X_party_nona = replace_na(X_party, stat_mode(X_party)),
    X_party_missing = if_else(is.na(X_party), 1, 0),
    X_ideology_nona = replace_na(X_ideology, median(X_ideology, na.rm = TRUE)),
    X_ideology_missing = if_else(is.na(X_ideology), 1, 0)
  )
```

`check_missingness_nona()` then confirms the imputation took. `has_nona_version`
says a companion column exists and `nona_has_na` says whether it still contains
missing values, which it should not.

```{r check-nona}
check_missingness_nona(dat, study_id = "demo_study_1")
```

## Did randomization balance the covariates?

`check_balance()` runs one test per covariate and one joint test of all of them
together. It returns a two-element list; `flatten = TRUE` returns a single tibble
instead, distinguished by a `test` column.

Because assignment was clustered, `clusters` and `se_type` are passed through to
the underlying `lm_robust()` call. This raises a warning, and the warning is
worth reading rather than suppressing.

```{r balance, warning = TRUE}
balance <- check_balance(
  dat, Z,
  covariates = c("X_age_nona", "X_income_nona", "X_ideology_nona",
                 "X_gender", "X_party_nona"),
  clusters = cluster_id,
  se_type = "CR2",
  study_id = "demo_study_1"
)

balance$covariate_tests
```

`F_stat` is the omnibus statistic from regressing the covariate on treatment and
`p_value` is its p-value. There is no coefficient column: with three or more arms
there would be no single coefficient to report, so the family reports the omnibus
test throughout. A factor covariate contributes one row per non-reference level,
which is why `X_party_nona` appears twice.

The joint test is the one the warning was about.

```{r balance-joint}
balance$joint_test
```

Under cluster randomization that p-value is not trustworthy. The parametric joint
test rejects roughly 13 percent of the time at a nominal 5 percent with 30
clusters, and no correction to its degrees of freedom repairs it, because the
cluster-robust variance estimator is itself biased downward when treatment is
constant within a cluster. Passing a `declaration` replaces the parametric
reference distribution with the randomization distribution, which is calibrated:

```{r balance-ri, eval = requireNamespace("ri2", quietly = TRUE)}
declaration <- declare_ra(clusters = dat$cluster_id)

check_balance(
  dat, Z,
  covariates = c("X_age_nona", "X_income_nona", "X_ideology_nona",
                 "X_gender", "X_party_nona"),
  declaration = declaration,
  sims = 200,
  study_id = "demo_study_1"
)$joint_test
```

The covariate-by-covariate tests need no such repair here. Treatment is binary, so
each is a single-degree-of-freedom test, and those are well calibrated with
cluster-robust standard errors. That is a fact about this design rather than about
covariate tests in general: with many arms and a small sample they over-reject just
as the joint test does.

`vignette("balance_testing")` works through the calibration of all of these.

### A p-value is not a magnitude

A significant imbalance in a large sample can be trivially small, and a
substantial one in a small sample can miss significance. `check_smd()` answers
the question the p-value cannot, reporting each difference in units of the
reference arm's standard deviation.

```{r smd}
check_smd(dat, Z,
          covariates = c("X_age_nona", "X_income_nona", "X_ideology_nona"),
          study_id = "demo_study_1") |>
  select(covariate, arm, reference, smd, flag)
```

`flag` marks an absolute SMD above 0.1, a conventional threshold. Check the
`reference` column rather than assuming it: the default is the first level, which
is a guess based on ordering, and if your control arm is not first then every sign
is inverted. Pass `reference` explicitly when in doubt. If the balance tests were
weighted, pass the same `weights` column here too, since an unweighted SMD beside
a weighted p-value describes a different sample.

## Did treatment affect who is missing an outcome?

`check_attrition()` regresses a missingness indicator on treatment for each
outcome. `Y_attitude` was built with differential attrition and `Y_behavior` was
not.

```{r attrition}
check_attrition(dat, Z, clusters = cluster_id, se_type = "CR2",
                study_id = "demo_study_1")
```

`Y_attitude` is flagged and `Y_behavior` is not, which is what the simulation put
there. This covariate-free test is well calibrated under clustering, so unlike the
joint balance test it needs no randomization-inference counterpart.

The `estimable` column matters when these results are stacked. An outcome nobody
skipped supports no test at all, and such a row reports `NA` rather than a p-value
of 1. A collection of studies contains many such outcomes, and counting them as
p-values near 1 would make the p-value diagnostics in
`vignette("checking_many_studies")` report a badly non-uniform distribution when
nothing is wrong.

Passing `covariates` fits the Lin (2013) interacted model instead, testing
treatment and all treatment-by-covariate interactions jointly. That asks a more
demanding question, whether attrition differs by treatment *anywhere* in covariate
space, and it spends a degree of freedom per covariate per arm to ask it. Read the
two together rather than replacing one with the other.

### Is covariate missingness itself balanced?

Imputation assumes covariate missingness is unrelated to treatment. Since the
`_missing` indicators were created above, that assumption is directly testable.

```{r missing-balance, warning = FALSE}
check_balance(
  dat, Z,
  covariates = c("X_income_missing", "X_party_missing"),
  clusters = cluster_id,
  se_type = "CR2",
  study_id = "demo_study_1"
)$covariate_tests
```

Balance here supports the imputation. Imbalance would suggest missingness is
related to treatment, which imputing a median cannot fix.

## What to do with the results

Three of the four checks are about cleaning and one is about the design.
Out-of-bounds outcomes, covariates missing an imputed companion, and companions
that still contain `NA` are all errors to fix at the source. Balance and attrition
results are different: they are evidence to report, not thresholds to act on.

- **A single flagged balance test** is usually not worth acting on. Under a valid
  design, roughly `alpha` of these tests reject by construction.
- **Differential attrition** is worth reporting as a limitation, and worth
  bounding or weighting if it is large.
- **Imbalanced missingness indicators** point at missingness that is not at
  random, where a sensitivity analysis is more honest than an imputation.

Dropping studies on the strength of a flagged check is how you introduce the bias
the check was meant to detect. `vignette("checking_many_studies")` makes that case
with the arithmetic behind it, and shows how to run these checks across a corpus
and read the result.
