---
title: "Checking Many Studies at Once"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Checking Many Studies at Once}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 3.2
)
```

```{r setup, message = FALSE}
library(excheckr)
```

A meta-analysis or a multi-study project runs the same handful of design checks
on every study it includes, and then has to make sense of several hundred test
results at once. This vignette covers that second half: labelling per-study
output so it can be stacked, collapsing it into one object, and reading the
result without fooling yourself.

The pattern has three stages.

1. One checking script per study, writing its results to disk.
2. `stack_checks()` to bind them into a single object.
3. `report_checks()` and the p-value diagnostics to decide what needs attention.

## Some studies to check

Real projects read cleaned data from disk. Here we simulate eight studies that
follow the column conventions the package expects: a treatment `Z`, covariates
prefixed `X_` with imputed `_nona` companions, and outcomes prefixed `Y_` scaled
to the unit interval.

Two studies are given deliberate problems, so there is something to find.

```{r simulate}
set.seed(20260725)

simulate_study <- function(study_id, n = 800, imbalanced = FALSE,
                           attrition = FALSE, out_of_bounds = FALSE) {
  X_age <- rnorm(n, 50, 15)
  X_educ <- factor(sample(c("HS", "College", "Postgrad"), n, replace = TRUE))

  # An imbalanced study assigns treatment as a function of a covariate
  Z <- if (imbalanced) {
    rbinom(n, 1, plogis((X_age - 50) / 12))
  } else {
    rbinom(n, 1, 0.5)
  }

  Y_support <- plogis(0.4 * Z + 0.02 * (X_age - 50) + rnorm(n))
  Y_turnout <- plogis(0.2 * Z + rnorm(n))

  # Differential attrition: the treated drop out more often
  p_missing <- if (attrition) 0.05 + 0.20 * Z else 0.08
  Y_support[rbinom(n, 1, p_missing) == 1] <- NA

  if (out_of_bounds) Y_turnout <- Y_turnout * 100

  # A covariate with missing values, plus its imputed companion
  X_age[sample(n, round(0.05 * n))] <- NA

  data.frame(
    study_id = study_id, Z = Z,
    X_age = X_age, X_age_nona = ifelse(is.na(X_age), mean(X_age, na.rm = TRUE), X_age),
    X_educ = X_educ, X_educ_nona = X_educ,
    Y_support = Y_support, Y_turnout = Y_turnout
  )
}

studies <- c(sprintf("study_%02d", 1:8))
datasets <- lapply(studies, function(s) {
  simulate_study(
    s,
    imbalanced    = s == "study_03",
    attrition     = s == "study_05",
    out_of_bounds = s == "study_07"
  )
})
names(datasets) <- studies
```

## One checking script per study

In a real project each study gets its own script, so that a study's covariate
list and quirks live next to the study rather than in a loop. The important
detail is `study_id`: every check takes it, and every check appends it to its
output, which is what makes the results stackable later.

`check_balance()` normally returns a two-element list. Passing `flatten = TRUE`
returns one tibble instead, with a `test` column distinguishing the joint test
from the covariate-by-covariate tests.

```{r checks}
check_dir <- file.path(tempdir(), "checks")
dir.create(check_dir, showWarnings = FALSE)

for (s in studies) {
  dat <- datasets[[s]]
  nona <- c("X_age_nona", "X_educ_nona")

  saveRDS(
    list(
      ybounds     = check_y_bounds(dat, study_id = s),
      missingness = check_missingness_nona(dat, study_id = s),
      balance     = check_balance(dat, Z, covariates = nona,
                                  study_id = s, flatten = TRUE),
      attrition   = check_attrition(dat, Z, outcomes = c("Y_support", "Y_turnout"),
                                    study_id = s)
    ),
    file.path(check_dir, paste0(s, "_checks.rds"))
  )
}

list.files(check_dir)
```

## Stacking

`stack_checks()` reads every per-study file and binds it element-wise, so the
`ybounds` tibbles from all eight studies become one `ybounds` tibble. Elements
that are `NULL` or have no rows in a given file are dropped before binding, so a
study that contributes nothing to one check does not break the others.

```{r stack}
all_checks <- stack_checks(check_dir)

names(all_checks)
nrow(all_checks$balance)
```

The stacked object is a plain named list of tibbles, so anything downstream is
ordinary data manipulation.

```{r stacked-peek}
head(all_checks$ybounds, 3)
```

## Triage

`report_checks()` applies the filters that a human would apply by hand and
returns only the rows that failed. Its print method shows the counts.

```{r report}
report <- report_checks(all_checks)
report
```

The three planted problems surface, and each element is a tibble you can inspect
directly.

```{r report-detail}
report$out_of_bounds[, c("study_id", "variable", "min", "max")]
report$attrition[, c("study_id", "outcome", "p_value")]
```

One column in the stacked attrition results is worth knowing about before the next
section. An outcome that nobody skipped supports no attrition test at all, and
those rows carry `estimable = FALSE` with an `NA` p-value rather than a p-value of
1. That distinction matters more than it looks: a real corpus contains many
outcomes with no attrition, and reporting them as ones would put a spike at the top
of the p-value distribution and make the diagnostics below announce a badly
non-uniform collection when nothing at all is wrong.

```{r estimable}
table(all_checks$attrition$estimable)
```

## Reading the p-values honestly

The flagged balance tests deserve more care than the other two categories, and
this is where multi-study checking most often goes wrong.

Under a valid design, balance test p-values are distributed Uniform(0, 1). That
means roughly `alpha` of them fall below `alpha` **by construction**. In a
collection of a few hundred tests, a handful of flags is what success looks
like. Treating each flagged test as evidence against the study that produced it,
and dropping those studies, is a reliable way to introduce exactly the bias the
checks were meant to detect.

The useful question is therefore not "which tests failed" but "does the
distribution look uniform". `summarize_check_pvalues()` answers it.

```{r summary}
covariate_tests <- all_checks$balance[all_checks$balance$test == "covariate", ]

summarize_check_pvalues(covariate_tests)
```

`pct_below` is the observed rejection rate and `expected_below` is what the
design implies, so the two should be close. `pct_below_fdr` applies a
Benjamini-Hochberg correction, which is the right lens when you want to know
whether any individual test survives multiplicity.

Two columns deserve more suspicion than the rest. `n_dropped` counts tests that
could not be computed and were excluded. When it is large relative to `n_tests`,
the summary describes a subsample selected on estimability rather than the corpus,
and says nothing at all about the studies that fell out. This is not a hypothetical:
the joint balance test relies on `nnet::multinom`, which fails to converge on small
strata, and in one real corpus 802 of 1143 joint balance tests were unestimable
while the summary reported `n_tests = 341` without comment.

`ks_p` tests the whole distribution against the uniform, but it assumes the
p-values are independent and these are not. Within a study, the indicators of one
factor covariate are mechanically dependent, correlated covariates add more, and
the joint test is a function of all of them. Read a small `ks_p` on a stacked
collection as evidence of dependence first and of a design problem second. The
`pct_below` against `expected_below` comparison is the more robust headline.

The `group` argument controls what the FDR correction is computed over. Adjusting
within study asks whether a given study looks broken; adjusting across all tests
asks whether the collection as a whole does. They answer different questions, so
the choice belongs at the call site rather than in a default.

```{r summary-grouped}
summarize_check_pvalues(covariate_tests, group = "study_id")
```

The picture is easier to read as a histogram against the uniform reference. The
dotted line is the expected height of each bar, and the dashed line marks
`alpha`.

```{r plot, message = FALSE}
plot_check_pvalues(covariate_tests)
```

Setting `fdr = TRUE` adds the adjusted panel alongside the raw one.

```{r plot-fdr, message = FALSE, fig.height = 3.4}
plot_check_pvalues(covariate_tests, group = "study_id", fdr = TRUE)
```

Because `plot_check_pvalues()` returns an ordinary `ggplot` object with minimal
theming, a project theme can be added to it in the usual way.

## Magnitude, not just significance

A p-value answers whether an imbalance is larger than chance would produce. It
does not answer whether the imbalance is large enough to matter, and in a large
sample a trivial difference will be significant. `check_smd()` reports the
standardized mean difference for each covariate against a reference arm,
dividing by the reference arm's standard deviation so the denominator is fixed
across arms.

```{r smd}
smd <- check_smd(datasets[["study_03"]], Z,
                 covariates = c("X_age_nona", "X_educ_nona"),
                 study_id = "study_03")

smd[, c("covariate", "level", "smd", "flag")]
```

This is the imbalanced study, and `X_age_nona` is flagged at the conventional
0.1 threshold. Running the same call on a sound study returns small values,
which is the comparison worth making.

```{r smd-clean}
check_smd(datasets[["study_01"]], Z,
          covariates = c("X_age_nona", "X_educ_nona"))[, c("covariate", "smd", "flag")]
```

## Where this leaves you

The output of this workflow is a decision about which studies need a closer
look, not an automatic exclusion rule. Out-of-bounds outcomes and covariates
missing an imputed companion are usually cleaning errors and worth fixing at
source. A single flagged balance test usually is not worth acting on. A
distribution of balance p-values that is visibly non-uniform is.

For the calibration behind the balance tests themselves, including how the joint
test behaves as the number of arms grows and when randomization inference is
worth the trouble, see
`vignette("balance_testing", package = "excheckr")`.
